### 2019-07-08 bp_conclusion.tex
adapting to the 2020 template

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness and focus at work are highly relevant aspects when it comes to productivity and well-being at the workplace. In this paper, we presented the results of a study with 14 professional knowledge workers in their workplace over an eight week period to examine the ability of biometrics for predicting these producitivity-related aspects. The longitude and in situ placement of the study---two key components of the study---as well as the breadth of human aspects examined, support and extend previous work. Based on twice collected daily survey responses and continually collected biometrics data we were able to create a model that was able to predict awakeness, stress, and focus, outperforming the baseline by as much as 84.6\% in the case of awakeness. The analysis further shows that there is a continuous increase in performance with the number of training instances, suggesting that there is more potential to improve the model's performance with further data points. Interestingly, while aspects such as stress have previously been predominantly linked to biometrics, our comparison with a model build from computer interaction features shows that even these types of features might bear big potential in predicting some of these productivity-related human aspects in the workplace.

>>>>

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness and focus at work are highly relevant aspects when it comes to productivity and well-being at the workplace. In this paper, we presented the results of a study with 14 professional knowledge workers in their workplace over an eight week period to examine the ability of biometrics for predicting these producitivity-related aspects. The longitude and in situ placement of the study---two key components of the study---as well as the breadth of human aspects examined, support and extend previous work. Based on twice collected daily survey responses and continually collected biometrics data we were able to create a model that was able to predict awakeness, stress, and focus, outperforming the baseline by as much as 84.6\% in the case of awakeness. The analysis further shows that there is a continuous increase in the model performance with the number of training instances, suggesting that there is a lot more potential to improve the model's performance with further data points. Interestingly, while aspects such as stress have previously been predominantly linked to biometrics, our comparison with a model build from computer interaction features shows that even these types of features might bear big potential in predicting some of these productivity-related human aspects in the workplace.

### 2019-07-08 bp_data.tex
adapting to the 2020 template

Reasons for a survey response to lack associated biometric data include:
\begin{itemize}
\item The participant not wearing the Everion in the hour before before beginning the survey
\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
\item Data not being uploaded successfully to the server
%\item Compatibility issues between the sensor and the OS tracking the participant's data
%\item Issues accessing the data uploaded by the participants
\end{itemize}

@@

\begin{figure}
  \centering
      \includegraphics[width=0.45\textwidth]{DuringTheDay.png}
  \caption{Availability of biometric data per participant. Green: survey responses with biometric
  data; red: responses with no biometric data.}
   \label{surveyBio}
   \vspace*{-6mm}
\end{figure}

@@

\begin{table}
  \centering
      \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
  \caption{Distribution of responses per participant over the 5-point Likert scale (1: not at all; 5: extremely) for each question. }
   \label{responseDistribution}
   \vspace*{-6mm}
\end{table}

>>>>

Reasons for a survey response to lack associated biometric data include:
\begin{itemize}
\item The participant not wearing the Everion in the hour before before beginning the survey
\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
\item Biometric data not being uploaded successfully to the server
%\item Compatibility issues between the sensor and the OS tracking the participant's data
%\item Issues accessing the data uploaded by the participants
\end{itemize}

@@

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{DuringTheDay.png}
  \caption{The figure shows the biometric data available per each participant. Green sections represent survey responses with biometric
  data available, red sections responses with no biometric data available}
   \label{surveyBio}
   \vspace*{-4mm}
\end{figure}

@@

\begin{table}
  \centering
      \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
  \caption{The distribution of the responses of each participant to the three questions asked during the day are shown. Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
   \label{responseDistribution}
   \vspace*{-9mm}
\end{table}

### 2019-07-08 bp_discussion.tex
adapting to the 2020 template

%\vspace{0.05in}
%\noindent
%\textbf{Corpus Size:}
%In this study we collected data from 14 professionals over the course of eight weeks. By scholarly standards this constitutes a large dataset --- especially given that the participants were not students. However, in the context of machine learning this corpus is relatively small. Thus, the model's overall performance, while promising, is only an indication of the performance that a larger corpus could provide. Our results show that there is a positive correlation between data added to the training corpus and the performance level of the model. This upwards trend likely continues, but a larger dataset is needed to confirm or reject this hypothesis. 
%
%
%\vspace{0.05in}
%\noindent
%\textbf{Per-User Model vs. Across-Users Model:}
%Our results show that individually trained models outperform on average a more general model. Yet, given the increase in model performance with training data set size, we expect that with enough data, a generic model will perform reasonably close to the individually trained models. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), and thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end.
%
%
%% \vspace{0.05in}
%% \noindent
%% \textbf{Per-User Model vs. Across-Users Model:}\
%% While we do see variations across participants, and thus our work shows that personal 
%% models perform better than generic models, we expect that with enough data, a generic 
%% model may perform equally well. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end and on evaluating the performance of a generic model.
%
%% \vspace{0.05in}
%% \noindent
%% \textbf{Data Collection:} 
%% One advantage of our study is that it was conducted on professionals in a real office setting (i.e., not on students). However, the disadvantage to working with professionals is that business deadlines and other pressing real world issues lead to missing survey responses. Additionally, busy professionals often fall into the trap of always selecting the same value for a given question that they have answered many times, which reduces the accuracy of the collected data. To minimize the effect of this inertia, we advise collecting data at typical low-productivity times, such as in the early afternoon or on Mondays~\cite{mark2014bored}.

>>>>

\vspace{0.05in}
\noindent
\textbf{Corpus Size:}
In this study we collected data from 14 professionals over the course of eight weeks. By scholarly standards this constitutes a large dataset --- especially given that the participants were not students. However, in the context of machine learning this corpus is relatively small. Thus, the model's overall performance, while promising, is only an indication of the performance that a larger corpus could provide. Our results show that there is a positive correlation between data added to the training corpus and the performance level of the model. This upwards trend likely continues, but a larger dataset is needed to confirm or reject this hypothesis.

% \vspace{0.05in}
% \noindent
% \textbf{Data Collection:} 
% One advantage of our study is that it was conducted on professionals in a real office setting (i.e., not on students). However, the disadvantage to working with professionals is that business deadlines and other pressing real world issues lead to missing survey responses. Additionally, busy professionals often fall into the trap of always selecting the same value for a given question that they have answered many times, which reduces the accuracy of the collected data. To minimize the effect of this inertia, we advise collecting data at typical low-productivity times, such as in the early afternoon or on Mondays~\cite{mark2014bored}.

\vspace{0.05in}
\noindent
\textbf{Per-User Model vs. Across-Users Model:}
Our results show that individually trained models outperform on average a more general model. Yet, given the increase in model performance with training data set size, we expect that with enough data, a generic model will perform reasonably close to the individually trained models. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), and thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end.

% \vspace{0.05in}
% \noindent
% \textbf{Per-User Model vs. Across-Users Model:}\
% While we do see variations across participants, and thus our work shows that personal 
% models perform better than generic models, we expect that with enough data, a generic 
% model may perform equally well. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end and on evaluating the performance of a generic model.

### 2019-07-08 bp_results_1.tex
adapting to the 2020 template

\begin{table*}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
  \caption{Prediction results individual models. The baseline row states averaged results of our baseline classifier, the general row states the averaged results of model trained on all participants. Distribution presents the distribution of the binarized outcome variable.}\label{tab:accuracy}%\todo{revise caption}}
%  The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables. The baseline row represents the averaged results of our baesline classifier. The general row shows the averaged results of our models trained on all participants.
  \vspace*{-4mm}
\end{table*}

@@

\subsection{Predicting Stress, Focus \& Awakeness}
% random forest best
\label{secOverallAccuracy}
As a first step, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performing a grid search analysis to determine the optimal hyperparameters for each classifier. Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, random forest, and a multilayer perceptron neural network. The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}. For the remainder of this paper, we will be referring to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
\begin{table}[t!]
	\begin{centering}
    \begin{tabular}{llll}
      \hline
      Variable & K & \# Estimators & Minimum Samples Split \\
      \hline
      Stress & 300 & 100 & 4\\
      Focus & 200 & 50 & 4\\
      Awakeness & 800 & 100 & 4\\
      \hline
    \end{tabular}
    \caption{Hyperparameters selected by grid search analysis for tuning our random forest models. K refers to the number of features selected.}\label{tab:hyperparams}
    \vspace*{-1cm}
    \end{centering}
\end{table}

@@

\begin{table}[t!]
  \begin{centering}
  \begin{tabular}{llll}
    \hline
    Feature Category & Stress & Focus & Awakeness\\
    \hline
    Heart Rate Variability & 18.3\% & 13\% & 13.6\%\\
    Blood Pulse Wave & 10\% & 14.2\% & 13.1\%\\
    Heart Rate & 8.7\% & 12.6\% & 10.3\%\\
    Skin Temperature & 15.7\% & 9.8\% & 10\%\\
    Galvanic Skin Response & 6.6\% & 8.2\% & 5.1\%\\
    Respiration Rate & 14.8\% & 12.7\% & 10\%\\
    Oxygen Saturation & 5.6\% & 4.3\% & 2\%\\ 
    Energy Expenditure & 6\% & 7.7\% & 4.8\%\\
    Activity & 4.6\% & 7.8\% & 7.6\%\\
    Steps & 1.7\% & 0.8\% & 0.9\%\\
    Time of Day & 0\% & 0.1\% & 0.5\%\\
    \hline
  \end{tabular}
  \caption{Averaged Gini importance of each feature category per response variable.}
  \label{tab:featureImportance}
  \end{centering}
  \vspace*{-7mm}
\end{table}

>>>>

\begin{table*}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
  \caption{Results of predictions using the individual models. The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables. The baseline row represents the averaged results of our baesline classifier. The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
  \vspace*{-4mm}
\end{table*}

@@

\subsection{Predicting Stress, Focus \& Awakeness}
% random forest best
\label{secOverallAccuracy}
As a first step, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performing a grid search analysis to determine the optimal hyperparameters for each classifier. Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, random forest, and a multilayer perceptron neural network. The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}. For the remainder of this paper, we will be referring to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
\begin{table}[h]
	\begin{centering}
    \begin{tabular}{llll}
      \hline
      Variable & K & \# Estimators & Minimum Samples Split \\
      \hline
      Stress & 300 & 100 & 4\\
      Focus & 200 & 50 & 4\\
      Awakeness & 800 & 100 & 4\\
      \hline
    \end{tabular}
    \caption{The hyperparameters selected by grid search analysis to tune our random forest models. K refers to the number of features selected.}    \label{tab:hyperparams}
    \end{centering}
\end{table}

@@

\begin{table}[h!]
  \begin{centering}
  \begin{tabular}{llll}
    \hline
    Feature Category & Stress & Focus & Awakeness\\
    \hline
    Heart Rate Variability & 18.3\% & 13\% & 13.6\%\\
    Blood Pulse Wave & 10\% & 14.2\% & 13.1\%\\
    Heart Rate & 8.7\% & 12.6\% & 10.3\%\\
    Skin Temperature & 15.7\% & 9.8\% & 10\%\\
    Galvanic Skin Response & 6.6\% & 8.2\% & 5.1\%\\
    Respiration Rate & 14.8\% & 12.7\% & 10\%\\
    Oxygen Saturation & 5.6\% & 4.3\% & 2\%\\ 
    Energy Expenditure & 6\% & 7.7\% & 4.8\%\\
    Activity & 4.6\% & 7.8\% & 7.6\%\\
    Steps & 1.7\% & 0.8\% & 0.9\%\\
    Time of Day & 0\% & 0.1\% & 0.5\%\\
    \hline
  \end{tabular}
  \caption{The averaged Gini importance of each feature category, per response variable.}
  \label{tab:featureImportance}
  \end{centering}
  \vspace*{-5mm}
\end{table}

### 2019-07-08 bp_results_2.tex
adapting to the 2020 template

\subsection{Minimum Time Window}\label{secMinimumTW}
In general, the less biometric data is needed to accurately predict a certain outcome measure, the easier and faster the analysis and data collection. To examine the optimal and minimum time window for the prediction of stress, focus, and awakeness, we used 16 different time windows from 10 seconds to 3 hours.% as depicted in Figure~\ref{timeWindows}. 
For our analysis, we then trained individual classifiers for each of the 16 time windows, using only features that had a time window smaller or equal to the time window rather than using all combinations of \small{$\{Biometric Measures\} X \{Statistical Metrics\} X \{Time Windows\}$}. We again used random forest and a leave-one-out cross validation to train individual classifiers. Since the number of features used for the training changed with each time window, we did not apply our feature selection in this case, but used all features available. Finally, due to the imbalance in the data, we again weighted each participants' classifier performance by the number of instances in the smaller class to calculate the average.

%\begin{figure}
%  \centering
%      \includegraphics[width=0.4\textwidth]{timeWindows.png}
%  \caption{Time windows of biometric data collected prior to each survey response: 10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, and 3hour.}
%   \label{timeWindows}
%\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabularx}{\columnwidth}{l l}
\hline
Feature collected by tool & Description \\ 
\hline
Total keystrokes per min& Sum of all types of keystrokes \\ 
Normal keystrokes per min&F[h] Not backspace and navigation \\ 
Backspace keystrokes per min& Backspace keystrokes \\ 
Navigation keystrokes per min& Arrow key keystrokes \\ 
Total clicks per min& Sum of all click types \\ 
Other clicks per min& Not right and left clicks \\ 
Left clicks per min& Left clicks \\ 
Right clicks per min& Right clicks \\ 
Scrolled distance per min& Scrolled distance in pixels \\ 
Moved distance per min& Mouse movements in pixels \\ 
Activity switches per min& Browser window title changes \\ 
Category switches per min& Activity performed category \\ 
\hline
\end{tabularx}
\caption{List of computer interaction features.}%~\cite{meyer18}}
\label{tracker}
\end{center}
\vspace*{-7mm}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabularx}{\columnwidth}{llllll}
\hline
Model/Feature Set & Accuracy & Precision & Recall & F-Score\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}Biometrics only & 0.808 & 0.269 & 0.314 & 0.289\\
\hspace{3mm}CI  & 0.758 & 0.425 & 0.362 & 0.391\\
\hspace{3mm}Biometrics + CI & 0.791 & 0.390 & 0.404 & 0.400\\
\hline
\textbf{Stress}\\
\hspace{3mm}Biometrics only & 0.775 & 0.270 &	0.260 & 0.265\\
\hspace{3mm}CI  & 0.698 & 0.290 & 0.272 & 0.281\\
\hspace{3mm}Biometrics + CI & 0.712 & 0.317 & 0.286 & 0.301\\
\hline
\textbf{Focus}\\
\hspace{3mm}Biometrics only & 0.716 & 0.251 & 0.256 & 0.253\\
\hspace{3mm}CI  & 0.742 & 0.332 & 0.342 & 0.337\\
\hspace{3mm}Biometrics + CI & 0.745 & 0.340 & 0.316 & 0.328\\
\hline
\end{tabularx}
\caption{Comparison of prediction performance for the 3 different features sets for 10 participants. CI stands for computer interaction.  Precision/recall refer to prediction of more important classes, i.e. stressed, not awake, not focused.}
%Comparison of the performance of predicting stress, focus and awakeness using the 3 different feature sets for the 10 participants. The performance is calculated as the average of the performance of the individual classifiers. Computer Interactions is abbreviated as C.I. here for readability. Precision and recall refer to the prediction of the more important classes, i.e. `stressed', `not awake', `not focused'.}
\label{ciPerformance}
\end{center}
\vspace*{-7mm}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabularx}{\columnwidth}{lc}
\hline
Model/Feature Set & Number of Features Selected\\
\hline
\textbf{Stress}\\
\hspace{3mm}CI & 400\\
\hspace{3mm}Biometrics + CI & 800\\
\hline
\textbf{Focus}\\
\hspace{3mm}CI & 20\\
\hspace{3mm}Biometrics + CI & 300\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}CI & All\\
\hspace{3mm}Biometrics + CI & 50\\
\hline
\end{tabularx}
\caption{Optimal number of features used for each of the three models. CI: computer interaction}
%The optimal number of features we found to select for each of the model/feature set combinations. Computer interactions is abbreviated as C.I.}
\label{ciFeatureSelection}
\end{center}
\vspace*{-7mm}
\end{table}

>>>>

\subsection{Minimum Time Window}\label{secMinimumTW}
In general, the less biometric data is needed to accurately predict a certain outcome measure, the easier and faster the analysis and data collection. To examine the optimal and minimum time window for the prediction of stress, focus, and awakeness, we used 16 different time windows from 10 seconds to 3 hours as depicted in Figure~\ref{timeWindows}. For our analysis, we then trained individual classifiers for each of the 16 time windows, using only features that had a time window smaller or equal to the time window rather than using all combinations of $\{Biometric Measures\} X \{Statistical Metrics\} X \{Time Windows\}$. We again used random forest and a leave-one-out cross validation to train individual classifiers. Since the number of features used for the training changed with each time window, we did not apply our feature selection in this case, but used all features available. Finally, due to the imbalance in the data, we again weighted each participants' classifier performance by the number of instances in the smaller class to calculate the average.

\begin{figure}
  \centering
      \includegraphics[width=0.4\textwidth]{timeWindows.png}
  \caption{Time windows of biometric data collected prior to each survey response: 10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, and 3hour.}
   \label{timeWindows}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{l l}
\hline

Feature collected by tool & Description \\ 
\hline
Total keystrokes per min& Sum of all types of keystrokes \\ 
Normal keystrokes per min&F[h] Not backspace and navigation \\ 
Backspace keystrokes per min& Backspace keystrokes \\ 
Navigation keystrokes per min& Arrow key keystrokes \\ 
Total clicks per min& Sum of all click types \\ 
Other clicks per min& Not right and left clicks \\ 
Left clicks per min& Left clicks \\ 
Right clicks per min& Right clicks \\ 
Scrolled distance per min& Scrolled distance in pixels \\ 
Moved distance per min& Mouse movements in pixels \\ 
Activity switches per min& Browser window title changes \\ 
Category switches per min& Activity performed category \\ 
\hline
\end{tabular}
\caption{Features collected per user by the computer interaction tracker}%~\cite{meyer18}}
\label{tracker}
\end{center}
\vspace*{-4mm}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{llllll}
\hline
Model/Feature Set & Accuracy & Precision & Recall & F-Score\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}Biometrics only & 0.808 & 0.269 & 0.314 & 0.289\\
\hspace{3mm}C.I. only & 0.758 & 0.425 & 0.362 & 0.391\\
\hspace{3mm}Biometrics + C.I. & 0.791 & 0.390 & 0.404 & 0.400\\
\hline
\textbf{Stress}\\
\hspace{3mm}Biometrics only & 0.775 & 0.270 &	0.260 & 0.265\\
\hspace{3mm}C.I. only & 0.698 & 0.290 & 0.272 & 0.281\\
\hspace{3mm}Biometrics + C.I. & 0.712 & 0.317 & 0.286 & 0.301\\
\hline
\textbf{Focus}\\
\hspace{3mm}Biometrics only & 0.716 & 0.251 & 0.256 & 0.253\\
\hspace{3mm}C.I. only & 0.742 & 0.332 & 0.342 & 0.337\\
\hspace{3mm}Biometrics + C.I. & 0.745 & 0.340 & 0.316 & 0.328\\
\hline
\end{tabular}
\caption{Comparison of the performance of predicting stress, focus and awakeness using the 3 different feature sets for the 10 participants. The performance is calculated as the average of the performance of the individual classifiers. Computer Interactions is abbreviated as C.I. here for readability. Precision and recall refer to the prediction of the more important classes, i.e. `stressed', `not awake', `not focused'.}
\label{ciPerformance}
\end{center}
\vspace*{-6mm}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{lc}
\hline
Model/Feature Set & Number of Features Selected\\
\hline
\textbf{Stress}\\
\hspace{3mm}C.I. & 400\\
\hspace{3mm}Biometrics + C.I. & 800\\
\hline
\textbf{Focus}\\
\hspace{3mm}C.I. & 20\\
\hspace{3mm}Biometrics + C.I. & 300\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}C.I. & All\\
\hspace{3mm}Biometrics + C.I. & 50\\
\hline

\end{tabular}
\caption{The optimal number of features we found to select for each of the model/feature set combinations. Computer interactions is abbreviated as C.I.}
\label{ciFeatureSelection}
\end{center}
\vspace*{-7mm}
\end{table}

### 2019-07-08 bp_studyMethod.tex
adapting to the 2020 template

\vspace{0.5em}
\noindent\textbf{RQ4}: How do biometrics compare to computer interaction measurements for predicting stress, focus, and awakeness?

@@

%We answer RQ1 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
%machine learning models and by comparing the best performing models to baseline approaches.
%To answer RQ2, in Section~\ref{secLearningCurve} we identify the smallest number of training samples necessary to obtain satisfactorily high prediction accuracy and acceptably low mean squared error.
%We answer RQ3 in Section~\ref{secMinimumTW} by considering 16 different time windows for collection of biometric measurements and by identifying the smallest time window on which we can base our training features without sacrificing prediction accuracy.
%Finally, to answer RQ4, in Section~\ref{secCI} we compare the prediction accuracy of machine learning models trained on biometric data versus those trained on computer interaction data.

@@

Of the 14 participants, 11 were male, 3 female. The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+. The participants have an average number of years of professional experience of ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years. All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.

@@

\subsubsection{Biometric Sensors}
Figure~\ref{everion} illustrates Biovotion's Everion~\cite{everion}, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements. Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements.

\begin{figure}
  \centering
      \includegraphics[width=0.35\textwidth]{Everion.jpg}
  \caption{Everion sensor used to collect biometric measures.}
   \label{everion}\vspace*{-0.2cm}
\end{figure}

@@

\begin{table}[t!]
\begin{center}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability & Milliseconds (ms) \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data}
\label{signals}
\end{center}
\vspace*{-7mm}
\end{table}

@@

Following guidelines from similar previous studies~\cite{fogarty05,tanaka11}, we asked the participants to respond to each question using a 5-point Likert scale ranging from 1 (not at all) to 5 (extremely) awake/stressed/focused. Each participant response, as stored by Survey Gizmo, comprised the date, the time at which the response was initiated, the time at which the survey was submitted, the unique identifier for the participant, and the responses submitted by the participant.

>>>>

\vspace{0.5em}
\noindent\textbf{RQ4}: How do biometric measurements compare to computer interaction measurements for predicting stress, focus, and awakeness?

@@

We answer RQ1 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
machine learning models and by comparing the best performing models to baseline approaches.
To answer RQ2, in Section~\ref{secLearningCurve} we identify the smallest number of training samples necessary to obtain satisfactorily high prediction accuracy and acceptably low mean squared error.
We answer RQ3 in Section~\ref{secMinimumTW} by considering 16 different time windows for collection of biometric measurements and by identifying the smallest time window on which we can base our training features without sacrificing prediction accuracy.
Finally, to answer RQ4, in Section~\ref{secCI} we compare the prediction accuracy of machine learning models trained on biometric data versus those trained on computer interaction data.

@@

Of the 14 participants, 11 are male and 3 are female. The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+. The participants have an average number of years of professional experience of ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years. All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.

@@

\subsubsection{Biometric Sensors}
Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/categories/201633909-Everion-Device} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements.

\begin{figure}
  \centering
      \includegraphics[width=0.4\textwidth]{Everion.jpg}
  \caption{We used Biovotion's Everion to collect biometric measurements from the participants}
   \label{everion}
\end{figure}

@@

\begin{table}[h!]
\begin{center}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability & Milliseconds (ms) \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data}
\label{signals}
\end{center}
\vspace*{-4mm}
\end{table}

@@

Following guidelines from similar previous studies~\cite{fogarty05,tanaka11}, we asked the participants to respond to each question using a 5-point Likert scale ranging from 1 (not at all awake/stressed/focused) to 5 (extremely awake/stressed/focused). Each participant response, as stored by Survey Gizmo, comprised the date, the time at which the response was initiated, the time at which the survey was submitted, the unique identifier for the participant, and the responses submitted by the participant.

### 2019-07-08 bp_threats.tex
adapting to the 2020 template

Another threat is that our results might not generalize
to a different office environment. We have conducted
this study in a typical office environment, common
among technology workers across the world.
These office environments control for a series of
variables to make them standard world wide such
as controlled temperature and lighting.

@@

%It is possible that the amount of data collected
%is not enough
%to draw valid conclusions. To address this threat, 
%we collected a data for an eight week period, which is
%400\% longer than previous studies~\cite{zuger18,Muller16}.

>>>>

Another threat is that our results won't generalize
to a different office environment. We have conducted
this study in a typical office environment, common
among technology workers across the world.
These office environments control for a series of
variables to make them standard world wide such
as controlled temperature and lighting.

@@

It is possible that the amount of data collected
is not enough
to draw valid conclusions. To address this threat, 
we collected a data for an eight week period, which is
400\% longer than previous studies~\cite{zuger18,Muller16}.
